---
title: "EDA"
author: "Anh Chi Tuan"
date: "2025-11-02"
output: html_document
---

# Preparation

```{r}
# Load library
library(sf) # spatial data, for geojson
library(dplyr) # filter data
library(corrplot) # For calculating the matrix
library(ggplot2) # For plotting the base heatmap
library(plotly)  # For interactivity
library(tidyr) # For data wrangling
library(glmnet)
library(Matrix) # Used by glmnet
library(rpart)
library(rpart.plot)
```

```{r}
# Read data
geojson_path <- "E:/Bayesian-Statistics-Project/data/merged_ecec_province.geojson"
my_sf <- read_sf(geojson_path)
```

```{r, echo=FALSE}
# Plot
plot(my_sf, max.plot=1)
```

```{r, echo=FALSE}
selected_provinces <- c("MI") 
selected_sf <- my_sf %>%
  filter(prov_sigla %in% selected_provinces)
plot(selected_sf, max.plot = 1)
```

```{r}
num_df <- my_sf %>% st_drop_geometry() %>%
  select(prov_sigla, prov_type, rip_code, reg_code, 16:last_col())
head(num_df)
```

??? There are many zero-values in the dataframe!??! Are these truly zeros or null?!

Assumption: These data are collected reasonably =\> There is no outlier, therefore, I don't deal with outliers.

# Correlation

```{r}
# 2. Calculate the Correlation Matrix
cor_matrix <- cor(num_df[5:ncol(num_df)], use = "pairwise.complete.obs")

cor_df <- as.data.frame(cor_matrix)
cor_df$Var1 <- rownames(cor_df)

# Use tidyr::pivot_longer to convert it to a long format
cor_long <- cor_df %>%
  tidyr::pivot_longer(cols = -Var1, 
                      names_to = "Var2", 
                      values_to = "Correlation") %>%
  # Filter to show only the upper triangle (optional, but good practice)
  filter(as.numeric(factor(Var1)) < as.numeric(factor(Var2)))


# --- 3. Create the Base Heatmap using ggplot2 ---

p <- ggplot(cor_long, aes(x = Var1, y = Var2, fill = Correlation, text = round(Correlation, 2))) +
  # Use geom_tile to create the color blocks
  geom_tile(color = "white", linewidth = 0.5) +
  
  # Add the correlation value as text in the center of the tile
  geom_text(aes(label = round(Correlation, 2)), size = 2) +
  
  # Set a divergent color scale (common for correlation)
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), name = "Correlation") +
  
  # Improve labels and theme
  labs(title = "Interactive Correlation Matrix", x = "", y = "") +
  theme_minimal() +
  # Rotate X axis labels for readability
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  # Enforce square aspect ratio
  coord_fixed()

# --- 4. Convert the ggplot object to an Interactive Plotly object ---

# Use the 'tooltip' argument to display the exact correlation value on hover
interactive_plot <- ggplotly(p, tooltip = c("Var1", "Var2", "text"))

# Display the interactive plot
interactive_plot
```

```{r}
# Set your desired correlation threshold (e.g., 0.7 for strong correlation)
threshold <- 0.7

# Convert the correlation matrix to a standard data frame
cor_df <- as.data.frame(cor_matrix)

# Use the column names as a vector of feature names
features <- colnames(cor_matrix)

# Find the indices of correlations >= threshold AND exclude self-correlations (i < j)
# We use the 'upper.tri' function to only look at the upper triangle of the matrix 
# (to avoid duplicates and self-correlations)
high_cor_indices <- which(upper.tri(cor_matrix, diag = FALSE) & (cor_matrix >= threshold), arr.ind = TRUE)

# --- 3. Extract and Display Pairs ---

# Extract the pairs and their correlation values
highly_correlated_pairs <- data.frame(
  Feature1 = features[high_cor_indices[, 1]],
  Feature2 = features[high_cor_indices[, 2]],
  Correlation = cor_matrix[high_cor_indices]
) %>%
  # Arrange by correlation value, descending
  arrange(desc(Correlation))

# Print the resulting table
cat(paste("--- Pairs with Correlation >= ", threshold, " ---\n"))
print(highly_correlated_pairs)
cat("---------------------------------------------------\n")
```

# Simple linear model

```{r}
# Try to fit with a linear model
lm_df <- num_df[3:ncol(num_df)]
lm_df$rip_code <- as.factor(lm_df$rip_code)
lm_df$reg_code <- as.factor(lm_df$reg_code)
lm_df$urban_type <- as.factor(lm_df$urban_type)
linear_model <- lm(fem_empl_rate ~ -1 + . + rip_code:reg_code, data = lm_df)
# summary(linear_model)

# Coefficients: (81 not defined because of singularities)
# => first, not consider reg_code and interaction rip_code and reg_code

linear_model <- lm(fem_empl_rate ~ -1 + . - reg_code, data = lm_df)
#summary(linear_model)
# => fem_house_rate can be removed
# 1. From model stats, p-value = 1.7e-09
# 2. Come back and check correlation, cor(inact_rate,fem_house_rate)>0.7

linear_model <- lm(fem_empl_rate ~ -1 + . - reg_code - fem_house_rate - empl_rate- unempl_rate - private_coverage, data = lm_df)
summary(linear_model)
```

```{r}
plot(linear_model)
```

The simple linear model seems reasonable to some extent (based on residual vs fitted and qq plot) (satisfies the hypothesis of normality)

# Linear model with Lasso

```{r}
formula <- as.formula(fem_empl_rate ~ -1 + . - reg_code)

# Y: Response Vector
Y <- lm_df$fem_empl_rate

# X: Predictor Matrix
# The -1 removes the intercept, aligning with the LASSO practice of standardizing predictors.
X <- model.matrix(formula, data = lm_df)[, -1]
```

```{r}
# Find Optimal Lambda (Tuning) using Cross-Validation
cv_lasso <- cv.glmnet(
  x = X, 
  y = Y, 
  alpha = 1,       # alpha = 1 specifies LASSO (L1 regularization)
  nfolds = 3      # Use 3-fold cross-validation
)

# Store the best lambda value found by cross-validation
# lambda.min gives the lambda with the smallest Mean Squared Error (MSE)
best_lambda <- cv_lasso$lambda.min
plot(cv_lasso)
```

```{r}
# Fit the final LASSO Model using the best lambda
lasso_model <- glmnet(
  x = X,
  y = Y,
  alpha = 1,
  lambda = best_lambda
)

# Extract the non-zero coefficients
lasso_coefficients <- coef(lasso_model)

# Convert to a data frame and filter for non-zero (selected) predictors
selected_predictors <- as.data.frame(as.matrix(lasso_coefficients)) %>%
  tibble::rownames_to_column(var = "Predictor") %>%
  rename(Coefficient = s0) %>%
  filter(Coefficient != 0)

cat("--- LASSO Selected Predictors and Coefficients (at Lambda =", round(best_lambda, 5), ") ---\n")
print(selected_predictors)
cat("----------------------------------------------------------------------------------------\n")
```

# Decision Tree

```{r}
dt_model <- rpart(
  formula,
  data = lm_df,
  method = "anova",
  control = rpart.control(cp = 0.001) # Optionally lower complexity parameter for a larger tree
)

# --- 5. Print the Text Summary and Model Details ---
print("--- Text Summary of Decision Tree Splits ---")
print(dt_model)

```

```{r}
rpart.plot(dt_model,
           type = 4,           # Draw the split labels closer to the split
           extra = 1,          # Display the number of observations at the node
           main = "Simple Regression Decision Tree for Female Employment Rate",
           box.palette = "Blues", 
           shadow.col = "gray"
           )
```

```{r}
dt_model$variable.importance
```
